A Bayesian statistical model is able to predict target-by-target selection behaviour in a human foraging task

### 2022-09-22 Analysis/index_revisions.Rmd
some initial work on revisions

```{r, message=FALSE}
# attach packages
library(tidyverse)
library(patchwork)
library(tidybayes)
library(ggrepel)
library(truncnorm)
library(modelr)
library(brms)
```

```{r}
# set ggplot2 options
theme_set(ggthemes::theme_tufte() + 
            theme(plot.background = element_rect(fill = "#fffff8", colour = "#fffff8")))

@@

Terminology questions:
- item v target
- participant v observer

@@

ggplot(a_cut, aes(b_bin, acc, colour = condition, fill = condition, shape = condition)) + 
  geom_point() + 
  geom_line(stat = "smooth", method = "loess", 
            formula = y ~ x, alpha = 0.65, size = 2, se = F) +
  geom_abline(linetype = 2) +
  scale_x_continuous("largest item weight") +
  scale_y_continuous("proportion of times selected") +
  theme(legend.position = "bottom") -> plt1

@@

#ggsave("../Figures/qjep_preds.pdf", 
#       plt_b + plt1 + plt2 + 
#         plot_layout(guides = "collect") & theme(legend.position = "bottom"),
#       width = 10, height = 4)

>>>>

```{r, message=FALSE, warning = FALSE}
# attach packages
library(tidyverse)
library(patchwork)
library(tidybayes)
library(ggrepel)
library(truncnorm)
library(modelr)
library(brms)
```

```{r}
# set ggplot2 options
#theme_set(ggthemes::theme_tufte() + 
#            theme(plot.background = element_rect(fill = "#fffff8", colour = "#fffff8")))

# for saving graphs for paper
theme_set(ggthemes::theme_tufte() + 
            theme(plot.background = element_rect(fill = "#ffffff", colour = "#ffffff")))

@@

@@

ggplot(a_cut, aes(b_bin, acc, colour = condition, fill = condition)) + 
  geom_point() + 
  geom_line(stat = "smooth", method = "loess", 
            formula = y ~ x, alpha = 0.65, size = 2, se = F) +
  geom_abline(linetype = 2) +
  scale_x_continuous("largest item weight") +
  scale_y_continuous("proportion of times selected") +
  theme(legend.position = "bottom") -> plt1

@@

# Making a slightly different version of plt_b for the revisions
ggplot(a_agg, aes(x = found, y = meanb, colour = condition, fill = condition)) + 
  geom_jitter(data = filter(a_agg, found<40), width = 0.1, height = 0, alpha = 0.2) + 
  stat_lineribbon(.width = 0.67, alpha = 0.50) +
  geom_path(data = filter(a_agg, observer == 1), 
                          aes(y = chance), linetype = 2, colour = "black") + 
  geom_point(data = tibble(x=40, y=1), aes(x, y), size = 1.5, colour = "black", fill = "grey") + 
  scale_x_continuous(breaks = c(2, 10, 20, 40), "target selection", expand = c(0.01, 0.01)) + 
  scale_y_continuous("average weight from model", expand = c(0.01, 0.01)) + facet_grid(~condition) +
  guides(color = "none", fill = "none") -> plt_b2

ggsave("../Figures/qjep_preds.pdf", 
       plt_b2 / (plt1 + plt2) + 
       plot_layout(guides = "collect") & theme(legend.position = 'bottom'),
       width = 10, height = 6)

### 2022-09-22 manuscriptRevisions.tex
some initial work on revisions

Foraging, the act of searching for and gathering multiple targets (such as food), has been studied in both non-human animal contexts \cite{dawkins1971} and in human psychology studies \cite{kristjansson2014}. Foraging engages a wide range of different perceptual, cognitive, and decision-related skills, and is an ecologically relevant behaviour for most species, including humans. For these and other reasons, a sustained interest in foraging has developed our understanding of how we search for multiple instances of multiple types of targets. For example, the classic marginal value theorem \cite {charnov1976} is generally good at predicting when an organism will decide to stop searching in a patch and move on to another, based on the finding rate dropping below an expected rate, and taking into account the relative energy costs of traveling between patches versus staying within a patch. There has also been a concerted effort to characterize the spatial patterns of foraging behaviour. The Lévy Walk, for example, has been argued to be a good description of the foraging path many species take when the locations of targets are unknown (e.g. \citep {bartumeus2005}, but see also \citep{benhamou2007}).

@@

To assess how well our model can account for behaviour at the level of individual trials, we start by stepping through each trial in the data from \cite{clarke2022}. For each target selection\footnote{Note: for the reasons discussed above, we ignore the initial target selection for now, as the model has very little to say about it. We return to this issue in section 4.}, in each trial, we can estimate how likely each of the remaining items are to be selected using the parameters from our posterior model. As can be seen in Figure \ref{fig:cal}, the model is putting somewhere between 25\% and 75\% of the weight on the target that is selected next, easily outperforming a chance $1/n$ baseline. We can also see  that the model is well calibrated in that the probabilities assigned to the largest target manage to capture how often that target is actually selected. However, we can see a decrease in accuracy when it comes to how often ``runner-up`` candidate items are selected as our model appears to systematically undervalue these (presumably by putting too much of the probabilistic weight on low chance items).

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/qjep_preds.pdf}
\caption{(\textit{left}) Posterior probabilities for target selections during a visual foraging task. Each dot shows the  data from an individual participant, averaged over trials, in a condition and the shaded region indicates the interval in which we expect 67\% of participants to fall. The dashed line indicates chance performance. (\textit{centre})  Calibration plot for our foraging model. The $x$-axis gives the largest weight assigned by the model while the $y$-axis shows how often that target was actually selected by a human participant. The dashed line here and in the right plot is the identity line. (\textit{right}) This plot shows how often the 2nd and 3rd ranked items are selected based on the weights assigned by the model.}
\label{fig:cal}
\end{figure}

@@

We now look at some example trials and compare the behaviour of our human participants to the predicted behaviour from our model (see Figure \ref{fig:qjep_paths}). When looking at trials in which our model has done a particularly good job of account for the target selections (model accuracy > 80$\%$) we can see that the \textit{disagreements} appear to often occur in cases when the human participant appears to be carrying out some form of local path length optimisation. In general, the cases in which our model suggests a different target from the one the participant actually selected appear reasonable.

@@

For trials in which we do a worse job of predicting (accuracy <50$\%$) we can see that the human behaviour appears less organised in general and it seems unlikely that adding some form of path-length optimisation to our model would improve accuracy with these participants. We can see this more clearly in Figure \ref{fig:qjep_path_lengths}: while our model does a good job of capturing the average total path length, there is considerable variation with some participants generating shorter paths and some longer. We can see  that this appears to be systematically related to how well our model can predict behaviour.

@@

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/init_sel_hex_plot.pdf}
\caption{Hexagonal heatmaps showing the two-dimensional distribution of the location of initial target selections. Grey areas indicate cells with a count of 0. (\textit{left}) shows the distribution over all initial target selections while (\textit{right}) shows the distribution of each participant's median initial selection. }
\label{fig:qjep_init_sel_hex}
\end{figure}

@@

However, we can also delve deeper into exploratory hypotheses about other factors that might influence predictability. One suggestion is that the people who our model is able to predict are in some way the most 'optimal'. If this is the case, we might expect that model predictability should correlate with how good people are at the task. There is some evidence that on a trial-by-trial basis, shorter trials are easier for the model to predict, and on a target-by-target basis, faster moves from one target to the next are more easily predicted by the model (see \textit{Supplementary Materials: Part 1}). This likely is linked to the fact that our model is better at predicting participants with a stronger proximity bias, as targets that are closer to the previous target are also likely to be selected more quickly. Intuitively, it makes sense that the model struggles to make predictions when a participant has exhausted a local patch and may need to make a bigger jump to a more distant area to carry on collecting targets, as there may be multiple candidates and distance may no longer be such a good predictor (for example, people may choose to move to the next most densely populated patch).

@@

Our exploratory analyses seem to hint at the idea that the people who are difficult for the model to predict are those who do not behave in a manner that optimises path length. It could be that these participants are simply unpredictable, and do not behave in a consistent manner on a target-by-target basis, making it difficult for a model to predict their behaviour at this level of granularity. However, it is possible that there are other factors that we have not taken into account that could further help explain these discrepancies. For example, it is possible that relatively unpredictable participants use multiple rules and switch between them in ways that we are not currently captured in the model. Some participants may also lose track of where they are more frequently (perhaps due to attentional lapses) and thus may show foraging patterns that are predictable apart from unpredictable discontinuities, which again may not be easy to model. Finally, the displays had artificial discontinuities (i.e. screen edges) which may lead to different behaviour in different participants: some may carry out the task as if they are reading, reaching the right hand edge and then swinging back to the left, whereas others may 'bounce' off an edge they have just reached. Our model may well be able to predict the latter, but could struggle with the former behaviour.

>>>>

Foraging, the act of searching for and gathering multiple targets (such as food), has been studied in both non-human animal contexts \cite{dawkins1971} and in human psychology studies \cite{kristjansson2014}. Foraging engages a wide range of different perceptual, cognitive, decision-related and motor skills, and is an ecologically relevant behaviour for most species, including humans. For these and other reasons, a sustained interest in foraging has developed our understanding of how we search for multiple instances of multiple types of targets. For example, the classic marginal value theorem \cite {charnov1976} is generally good at predicting when an organism will decide to stop searching in a patch and move on to another, based on the finding rate dropping below an expected rate, and taking into account the relative energy costs of traveling between patches versus staying within a patch. There has also been a concerted effort to characterize the spatial patterns of foraging behaviour. The Lévy Walk, for example, has been argued to be a good description of the foraging path many species take when the locations of targets are unknown (e.g. \citep {bartumeus2005}, but see also \citep{benhamou2007}).

@@

To assess how well our model can account for behaviour at the level of individual trials, we start by stepping through each trial in the data from \cite{clarke2022}. For each target selection\footnote{Note: for the reasons discussed above, we ignore the initial target selection for now, as the model has very little to say about it. We return to this issue in section 4.}, in each trial, we can estimate how likely each of the remaining items are to be selected using the parameters from our posterior model. As can be seen in Figure \ref{fig:cal}, the model is putting somewhere between 25\% and 75\% of the weight on the target that is selected next, easily outperforming a chance $1/n$ baseline. We can also see  that the model is well calibrated in that the probabilities assigned to the most likely target manage to capture how often that target is actually selected. However, we can see a decrease in accuracy when it comes to how often ``runner-up`` candidate items are selected as our model appears to systematically undervalue these (presumably by putting too much of the probabilistic weight on low chance items).

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/qjep_preds.pdf}
\caption{(\textit{top left}) Posterior probabilities for target selections during a visual foraging task. Each dot shows the data from an individual participant, averaged over trials, in the feature condition and the shaded region indicates the interval in which we expect 67\% of participants to fall. The dashed line indicates chance performance. (\textit{top right}) As top left, except showing the conjunction condition. (\textit{bottom left})  Calibration plot for our foraging model. The $x$-axis gives the largest weight assigned by the model while the $y$-axis shows how often that target was actually selected by a human participant. The dashed line here and in the right plot is the identity line. (\textit{bottom right}) This plot shows how often the model selects the 2nd and 3rd ranked items based on the weights assigned by the model.}
\label{fig:cal}
\end{figure}

@@

We now look at some example trials and compare the behaviour of our human participants to the predicted behaviour from our model (see Figure \ref{fig:qjep_paths}). When looking at trials in which our model has done a particularly good job of accounting for the target selections (model accuracy > 80$\%$) we can see that the \textit{disagreements} often occur in cases when the human participant appears to be carrying out some form of local path length optimisation. In general, the cases in which our model suggests a different target from the one the participant actually selected appear reasonable.

@@

For trials in which the model does a worse job of predicting (accuracy <50$\%$) we can see that the human behaviour appears less organised in general and it seems unlikely that adding some form of path-length optimisation to our model would improve accuracy with these participants. We can see this more clearly in Figure \ref{fig:qjep_path_lengths}: while our model does a good job of capturing the average total path length, there is considerable variation with some participants generating shorter paths and some longer. We can see  that this appears to be systematically related to how well our model can predict behaviour.

@@

\begin{figure}[H]
\centering
\includegraphics[width=12 cm]{Figures/init_sel_hex_plot.pdf}
\caption{Hexagonal heatmaps showing the two-dimensional distribution of the location of initial target selections. Grey areas indicate cells with a count of 0. (\textit{left}) shows the distribution over all initial target selections (i.e. multiple trials per participant) while (\textit{right}) shows the distribution of each participant's median initial selection (i.e. each participant contributes one data point to the graph). }
\label{fig:qjep_init_sel_hex}
\end{figure}

@@

However, we can also delve deeper into exploratory hypotheses about other factors that might influence predictability. One suggestion is that the people who our model is able to predict are in some way the most 'optimal'. If this is the case, we might expect that model predictability should correlate with how good people are at the task. There is some evidence that on a trial-by-trial basis, shorter trials are easier for the model to predict, and on a target-by-target basis, faster moves from one target to the next are more easily predicted by the model (see \textit{Supplementary Materials: Part 1}). This likely is linked to the fact that our model is better at predicting participants with a stronger proximity bias, as targets that are closer to the previous target are also likely to be selected more quickly. Intuitively, it makes sense that the model makes poorer predictions when a participant has exhausted a local patch and may need to make a bigger jump to a more distant area to carry on collecting targets, as there may be multiple candidates and distance may no longer be such a good predictor (for example, people may choose to move to the next most densely populated patch).

@@

Our exploratory analyses seem to hint at the idea that the people who are difficult for the model to predict are those who do not behave in a manner that optimises path length. It could be that these participants are simply unpredictable, and do not behave in a consistent manner on a target-by-target basis, making it difficult for a model to predict their behaviour at this level of granularity. However, it is possible that there are other factors that we have not taken into account that could further help explain these discrepancies. For example, it is possible that relatively unpredictable participants use multiple rules and switch between them in ways that we are not currently captured in the model. Some participants may also lose track of where they are more frequently (perhaps due to attentional lapses) and thus may show foraging patterns that are predictable apart from unpredictable discontinuities, which again may not be easy to model. Another possibility is that participants may make exploratory, information-seeking visits to different parts of the array due to uncertainty about the targets present: this may be particularly important for conjunction trials, where peripheral visual information about shape is likely to be poorer. Finally, the displays had artificial discontinuities (i.e. screen edges) which may lead to different behaviour in different participants: some may carry out the task as if they are reading, reaching the right hand edge and then swinging back to the left, whereas others may 'bounce' off an edge they have just reached. Our model may well be able to predict the latter, but make less good predictions for the former behaviour.
